---
title: "MIDGE Aim 2 Growth Curve Pipeline"
output: html_document
date: "2025-07-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tidyverse)
library(splines)
```

#Load in and merge gc_file and gc_layout
```{r}
data_folder <- "../data/growth_curves/"

# List all gc_file files
gc_files <- list.files(data_folder, pattern = "\\d{4}-\\d{2}-\\d{2}_gc_file.txt$", full.names = TRUE)

# Helper function to extract date from filename
get_date <- function(filename) {
  str_extract(filename, "\\d{4}-\\d{2}-\\d{2}")
}

process_experiment <- function(gc_file_path) {
  date <- get_date(gc_file_path)
  gc_layout_path <- file.path(data_folder, paste0(date, "_gc_layout.csv"))
  message("Processing: ", gc_file_path)
  
  if (!file.exists(gc_layout_path)) {
    warning("Matching gc_layout not found for ", gc_file_path, "; skipping.")
    return(NULL)
  }
  
  # Safely read and process the growth curve file
  gc_file <- tryCatch(
    read_tsv(gc_file_path, locale = locale(encoding = "latin1")) %>% select(-2),
    error = function(e) { warning("Could not read gc_file: ", gc_file_path); return(NULL) }
  )
  if (is.null(gc_file)) return(NULL)
  
  # Pivot to long format
  gc_file_long <- gc_file %>%
    pivot_longer(-Time, names_to = "well", values_to = "values")
  
  # Read and check the layout file
  gc_layout <- tryCatch(
    read_csv(gc_layout_path),
    error = function(e) { warning("Could not read gc_layout: ", gc_layout_path); return(NULL) }
  )
  if (is.null(gc_layout)) return(NULL)
  
  # Merge
  gc_merged <- left_join(gc_file_long, gc_layout, by = "well") %>%
    mutate(experimentdate = date)
  
  # Optionally: Print preview for each file
  print(head(gc_merged, 2))
  
  return(gc_merged)
}

# Combine all experiments
all_growth_data <- gc_files %>%
  map_dfr(process_experiment) %>%
  mutate(time = as.numeric(hms(Time)) / 3600 #time for hours
  ) %>%
  select(-Time) %>%
  rename(Time = "time") %>%
  select(Time, well, values, strain, experimentdate)

# Optional to inspect result
head(all_growth_data)
```

#adding technical and biological replicates means
```{r}
all_growth_data_noNA <- all_growth_data %>%
  na.omit() %>%
  group_by(strain, experimentdate, Time) %>%
  mutate(tech_avg = mean(values),
         tech_sd = sd(values),
         #Identify Outlier Wells per Group
         is_well_outlier = abs(values - tech_avg) > 2 * tech_sd)%>% # mark if 2 SDs from mean) 
   mutate(n_tech = n()) %>%
  ungroup() %>%
  group_by(strain, Time) %>%
  mutate(bio_avg = mean(tech_avg),
         bio_sd = sd(tech_avg),
         #Identify Outlier Experiments per Group
         is_bio_outlier = abs(tech_avg - bio_avg) > 2 * bio_sd) %>% # you can use 2 or 3 SDs) 
  mutate(n_bio = n_distinct(experimentdate)) %>%
  ungroup()
```

#individual curve of the technical replicates for each strain
```{r}
all_growth_data_noNA %>%
  ggplot(aes(x = Time, y = values, group = well, color = strain)) +
  geom_line(alpha = 0.6) +
  facet_wrap(strain ~ experimentdate) +
  labs(title = "Technical Replicates: Each Well per Strain and Date")
```

#individual curve for each of the biological replicate
```{r}
tech_summary <- all_growth_data_noNA %>%
  group_by(strain, experimentdate, Time) %>%
  summarize(tech_avg = mean(values), .groups = "drop")

ggplot(tech_summary, aes(x = Time, y = tech_avg, group = experimentdate, color = experimentdate)) +
  geom_line() +
  facet_wrap(~strain) +
  labs(title = "Biological Replicates: Mean Curve for Each Experiment Date per Strain")
```

#calculating the growth rate
##growth rate function
```{r}
library(splines)
get_slope <- function(x,y){
  # Fit smoothing spline
    spline_fit <- smooth.spline(x, log(y), spar = 0.6)

  # Predict derivative (slope)
  spline_deriv <- predict(spline_fit, deriv = 1)[[2]]
  
  return(spline_deriv)
}

concat_slope <- function(df, window_size) {
  # df <- analysis_data  %>%
  # mutate(value = ifelse(values<=0, 1e-7, values)) %>%
  # group_by(strain,well) %>%
  # arrange(strain,well,Time)
  # Safely return empty result if df is empty or too short
  if (nrow(df) < window_size) {
    return(data.frame(
      Time = numeric(0),
      OD = numeric(0),
      growth_rate = numeric(0),
      stringsAsFactors = FALSE
    ))
  }

  t <- df$Time
  od <- df$values

  summary_df <- data.frame(
    Time = numeric(0),
    OD = numeric(0),
    growth_rate = numeric(0),
    stringsAsFactors = FALSE
  )

  for(i in 1:(nrow(df) - (window_size - 1))) {
    x <- t[i:(i + window_size - 1)]
    y <- od[i:(i + window_size - 1)]
    slope <- get_slope(x = x, y = y)

    new_df <- data.frame(
      Time = t[i + round(window_size / 2) - 1],
      OD = od[i + round(window_size / 2) - 1],
      growth_rate = slope,
      stringsAsFactors = FALSE
    )

    summary_df <- rbind(summary_df, new_df)
  }

  return(summary_df)
}
  

```

##Calculating the growth rate 
```{r}
analysis_data <- all_growth_data_noNA 

growth_summary_df <- analysis_data  %>%
  mutate(value = ifelse(values<=0, 1e-7, values)) %>%
  group_by(strain,well) %>%
  arrange(strain,well,Time) %>%
  group_modify(~ concat_slope(.x, 4)) %>%
  ungroup() %>%
  group_by(strain,well,Time, OD) %>%
  summarise(growth_rate = mean(growth_rate, na.rm = TRUE), .groups = "drop") %>%
  ungroup()

growth_summary_df <- growth_summary_df %>%
  group_by(Time, strain, well) %>%
  merge(analysis_data, by = c("strain", "well", "Time"))
```

##Identifying the max growth rate
```{r}
###Check if the growth rate is consistent over time
growth_summary_df <- growth_summary_df %>%
  filter( OD>1e-02) %>%
  group_by(strain, experimentdate) %>%
  mutate(growth_rank  = rank(-growth_rate),
         similar_neighbor = ifelse(growth_rank - lead(growth_rank) <=abs(5) & growth_rank - lag(growth_rank) <=abs(5), TRUE, FALSE))

### Graph the growth rates
growth_summary_df %>%
  ggplot() + geom_point(aes(y = growth_rate, x = Time, color = strain)) + facet_wrap(.~strain) 
```

##Extract probably maximum growth rate
```{r}
max_growth_df <- growth_summary_df %>%
  group_by(strain, experimentdate) %>%
  #filter(OD>0.2) %>% # remove the OD values that are too low
  filter(similar_neighbor==TRUE)%>%
  filter(Time<=6) %>% # log phase
  filter(growth_rank ==min(growth_rank)) %>%
  group_by(strain) %>%
  mutate(avg_growth_rate = mean(growth_rate)) %>%
  ungroup()

# get the strains that has no growth
no_growth_df <- max_growth_df %>%
  filter(OD==0) %>%
  distinct(strain, experimentdate)


max_growth_df <- max_growth_df %>%
  mutate(OD_range = case_when(OD <= 0 ~ "OD = 0",
                   OD < 0.2 & OD>0 ~ "0 < OD < 0.2",
                   OD >= 0.2 ~ "OD>=0.2")) 


## plot the Condition and growth rates y: time
max_growth_df %>%
  ggplot(aes(x = growth_rate, y = Time, color = strain)) + geom_point(size=2) 

max_growth_df %>%
  filter(growth_rate!=0) %>%
  #filter(OD_range!="0 < OD < 0.2") %>%
  ggplot(aes(x = growth_rate, y = OD, color = strain)) + 
  geom_point(size =2) +
  geom_smooth(method = "lm", se = FALSE, color = "black")
```

##Identifying info on the trend line
```{r}
library(ggplot2)
library(broom)

# Fit linear model
model <- lm(OD ~ growth_rate, data = max_growth_df %>% filter(growth_rate != 0))

# Extract coefficients and R-squared
coefs <- coef(model)
r2 <- summary(model)$r.squared

# Create annotation string
eqn <- sprintf("y = %.3fx + %.3f\nR² = %.3f", coefs[2], coefs[1], r2)

max_growth_df %>%
  filter(growth_rate != 0) %>%
  ggplot(aes(x = growth_rate, y = OD)) +
  geom_point(aes(color = strain), size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  annotate("text", x = Inf, y = -Inf, label = eqn, hjust = 1.1, vjust = -0.4, size = 5, color = "black") +
  theme_minimal()
```

#Calculating the lag time



#Calculating the maxOD
```{r}
maxOD_df <- analysis_data %>%
  group_by(strain, experimentdate) %>%
  filter(values == max(values, na.rm = TRUE)) %>%
  ungroup() %>%
  rename(max_OD = values)

maxOD_df %>%
  ggplot(aes(x = strain, y = max_OD)) +
  geom_point(aes(color = strain), size = 2) +
  labs(title = "Maximum growth",
       y = "OD") +
  theme_minimal()
```

#Merging the dataframes
```{r}
summary_df <- max_growth_df %>%
  select(strain, growth_rate) %>%
  full_join(maxOD_df, by = "strain") %>%
  select(strain, growth_rate, max_OD)
  
```

